A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-06 podc_paper/algorithm-2.tex
typo
minor edits to GC code

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \State \gc{\var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\gc{    
	\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
	\State \var{max} \assign 0
	\For {$k \assign 1..p$}
		\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
	\EndFor 
	\State \Call{Help}{}
	\State \Call{Discard}{$\var{root}, \var{max}-2$}
	\EndIf
	\State
}

    \ForEach{\fld{dir} {\keywordfont{in}} \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHeadGC}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNullGC}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvanceGC}
        \EndIf
    \EndFor \label{endHelpChild1GC}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC} 
    \EndIf
    \State \gc{\var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
    \State \var{\Call{Advance}{\var{v, h}}}\label{advanceGC}
    \State \Return{ \var{result}}
\EndFunction{Refresh}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \gc{\var{new}.\fld{index} \assign \var{h}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
        \State \gc{\var{last[this process id]} \assign\ \var{b}}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
        \State \gc{\var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}}\label{findAnswerGC}
        \State \gc{\var{last[this process id]} \assign\ $b_e$}
        \State \gc{\Return{\var{e}}}
    \EndIf
\EndFunction{FindResponse}

@@

\Function{void}{Help}{}
    \State \linecomment{help pending dequeues that have been propagated to root}
    \For{\bf{each} $\ell$ in $\var{leaf}[1..k]$}
        \State $\var{h} \assign \ell.\head$
        \If{$\ell.\fld{blocks}[\var{h}].\fld{element} = \nl$ and \Call{Propagated}{$\ell, h$}}
            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, h, 1}$
            \State \var{output} \assign\ \Call{FindResponse}{$b, i$} 
            \State $\ell.\fld{blocks}[\var{h}].\fld{response} \assign\ \var{output}$
        \EndIf
    \EndFor
\EndFunction{Help}

@@

\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
    \State \linecomment{discard blocks with index less than or equal to threshold $t$ from RBT \var{v}.\fld{blocks} and all subblocks of those blocks}
    \State \var{t\sub{left}} = \var{v}.\fld{blocks}[\var{t}].\eleft
    \State \var{t\sub{right}} = \var{v}.\fld{blocks}[\var{t}].\eright
	\For{\var{j} \assign 1 to $p$}
        \State $T \assign \var{v}.\fld{blocks}$
        \State $T'  \assign \Call{Split}{T,t}$ \Comment{$T'$ is the root of new RBT containing $T$'s blocks with $\var{index} > t$}
        \State \bf{exit when} \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{splitCASGC}
    \EndFor
    \If{\var{v} is an internal \typ{Node}} \Comment{discard subblocks in descendants of $v$}
        \State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}
        \State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}
    \EndIf
\EndFunction{Discard}
}
\end{algorithmic}
\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
\end{figure}

If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.

>>>>

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

	\gc{\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}}
	\gc{\State \var{max} \assign 0}
	\gc{\For {$k \assign 1..p$}}
		\gc{\State $\var{max} \assign \max(\var{max}, \var{last}[k])$}
	\gc{\EndFor}
	\gc{\State \Call{Help}{}}
	\gc{\State \Call{Discard}{$\var{root}, \var{max}-1$}}
	\gc{\EndIf}

    \ForEach{\fld{dir} {\keywordfont{in}} \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHeadGC}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNullGC}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvanceGC}
        \EndIf
    \EndFor \label{endHelpChild1GC}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC} 
    \EndIf
    \gc{\State \var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
    \State \Return{ \var{result}}
\EndFunction{Refresh}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \gc{\State \var{new}.\fld{index} \assign \var{h}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
        \gc{\State \var{last}[\var{this process id}] \assign\ \var{b}}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
        \gc{
        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
        \State \var{last}[\var{this process id}] \assign\ $b_e$
        \State \Return{\var{e}}
        }
    \EndIf
\EndFunction{FindResponse}

@@

\Function{void}{Help}{}
    \State \linecomment{help pending dequeues that have been propagated to root}
    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{h} \assign \ell.\head$
        \If{$\ell.\fld{blocks}[\var{h}].\fld{element} = \nl$ and \Call{Propagated}{$\ell, h$}}
            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, h, 1}$
            \State \var{output} \assign\ \Call{FindResponse}{$b, i$} 
            \State $\ell.\fld{blocks}[\var{h}].\fld{response} \assign\ \var{output}$
        \EndIf
    \EndFor
\EndFunction{Help}

@@

\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
    \State \linecomment{discard blocks with index less than or equal to threshold $t$ from RBT \var{v}.\fld{blocks} and all subblocks of those blocks}
    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
    \State \var{t\sub{right}} \assign \var{v}.\fld{blocks}[\var{t}].\eright
	\For{\var{j} \assign 1 to $p$}
        \State $T \assign \var{v}.\fld{blocks}$
        \State $T'  \assign \Call{Split}{T,t}$ \Comment{$T'$ is the root of new RBT containing $T$'s blocks with $\var{index} \geq t$}
        \State \bf{exit when} \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{splitCASGC}
    \EndFor
    \If{\var{v} is an internal \typ{Node}} \Comment{discard subblocks in descendants of $v$}
        \State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}
        \State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}
    \EndIf
\EndFunction{Discard}
}
\end{algorithmic}
\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
\end{figure}

\begin{lemma}

\end{lemma}

-------
